Higher Moments Distances from Robust Subspace for Botnet Detection

### 2019-05-02 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
working on literature review (robust pca)

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unespected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomaly, and sometimes relies into visual analysis for pricipal component selection. Robust Principal Component Analysis (RPCA) \cite{candes2011robust} is an extensio of PCA that aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has receiving a growing atention of researchers aiming the development of network anomlay detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

@@

% botnet detection
A botnet is a network of compromised machines under the influence of malware (bot). The botnet is commandeered by a “botmaster” and utilized as “resource” or “platform” for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information exfiltration. The botmaster coordenate a botnet through a command and control (C\&C) channel where bots receive commands and syncronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and achitectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

Bhuyan \emph{et al.} \cite{acarali2016survey} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed Bhuyan \emph{et al.} \cite{acarali2016survey} present an analysis of four major categories of anomaly detection techniques which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), explains cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for detection.

Acarali \emph{et al.} \cite{acarali2016survey} surveys network-based detection approaches for HTTP-based botnets, surveys traffic-based features used to detect bot traffic and presents an abstraction of the main types of features in relation to protocols and OSI layers. Wang \emph{et al.} \cite{Wang2018ddosbotnetssurvey} present an analysis based on 50,704 different Internet DDoS attacks originated of 674 botnets from 23 different botnet families with a total of 9,026 victim belonging to 1,074 organizations in 186 countries. Their analysis reveals that geolocation of the attacking sources follows patterns and enables source prediction, and highlight that multiple attacks to the same target also exhibit strong patterns of inter-attack time interval, also presents that there is a trend for different botnets to launch DDoS attacks targeting the same victim, simultaneously or in turn.

@@

Khattak \emph{et al.} \cite{khattak2015botflex} proposed BotFlex, which is a network-based tool for botnet detection, based on a Complex Event Processing (CEP) engine and on a correlation framework that continuously receives events and correlates them according to rules. BotFlex's results are compared to BotHunter \cite{gu2007bothunter}, but the evaluation relies in a own and not public data set.

% justificative for the dataset and present who else use it
Several approaches for network attack detections uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection and proposes two methods (BClus and CAMNEP) and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The autors recommend that none of the botnet families used in the training and crossvalidation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training. In contrast, we evaluate each scenario of CTU-13 individually in a semi-supervised approach that does not rely on training data with anomalies.

@@

Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. Callegari \emph{et al.} \cite{callegari2011novel} propose a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of normal and anomalies according to principal components (normal) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} propose OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack.

Existing approaches [22] added a separated interpretation step to solve this problem by using techniques such as decision trees. However such indirect interpretation often failed to reveal the true causes of the anomalies detected by PCA-based methods [17]. Another recent work [11] proposed the joint sparse PCA (JSPCA) model to identify a low-dimensional approximation of the abnormal subspace, so that all anomalies can be localized onto a small subset of original feature variables. However, for individual anomaly interpretation, especially anomalies of different types, we need a more accurate and direct way of interpretation.

% Anomaly detection by subspace larning
Principal Component Analysis (PCA) is a statistical technique commonly used for dimensionality reduction. It uses an orthogonal transformation to convert a set of correlated variables into a set of linearly uncorrelated variables, where the first principal components have the largest variance. PCA has been used in attack detection \cite{almotairi2009technique}. However, PCA requires human intervention in order to identify abnormalities based on the eigenvalues profiles, if used without complementary techniques.

Use to justify the sensitivity of pca to outliers, only.  Additionally, Lee \emph{et al.} only evaluate their proposed scheme for covariance analysis, while we adopt an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for flood and probe attacks, respectively.

Besides being prone to higher errors and false positives, such human intervention makes PCA not useful for real time applications. Therefore, in order to automate the analysis of eigenvalues profile, model order selection (MOS) schemes should be incorporated. Vieira .... (The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probe and flood attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.)

% Anomaly detection by robust subspace larning
In contrast to classical PCA, our robust PCA detector does not require having a perfect ground-truth for training; it just trains from the (eventually contaminated) background traffic. It is, in fact, an unsupervised learning algorithm. Moreover, the addition of feature selection allows matching the specificities of the traffic the detector is operating on, which can vary widely both in terms of the most frequent attacks and the relative weight of licit applications. Feature selection does require training with a labeled data set bu

@@

@@

We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for a robust subspace learning

\cite{zhou2017anomaly} (Anomaly detection with robust deep autoencoders)

>>>>

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unespected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomaly, and sometimes relies into visual analysis for pricipal component selection. Robust Principal Component Analysis (RPCA) \cite{candes2011robust} is an extension of PCA that aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has receiving a growing atention of researchers aiming the development of network anomlay detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

@@

% network anomaly detectiom
Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), explains cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for detection.

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information exfiltration. We refer to \cite{ahmed2016survey,moustafa2019holistic} for an overview of network attacks. The botmaster coordenate a botnet through a command and control (C\&C) channel where bots receive commands and syncronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and achitectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

Acarali \emph{et al.} \cite{acarali2016survey} surveys network-based detection approaches for HTTP-based botnets, surveys traffic-based features used to detect bot traffic and presents an abstraction of the main types of features related to protocols and OSI layers. Wang \emph{et al.} \cite{Wang2018ddosbotnetssurvey} present an analysis based on 50,704 different Internet DDoS attacks originated of 674 botnets from 23 different botnet families with a total of 9,026 victim belonging to 1,074 organizations in 186 countries. Their analysis reveals that geolocation of the attacking sources follows patterns and enables source prediction, and highlight that multiple attacks to the same target also exhibit strong patterns of inter-attack time interval, also presents that there is a trend for different botnets to launch DDoS attacks targeting the same victim, simultaneously or in turn.

@@

Khattak \emph{et al.} \cite{khattak2015botflex} proposed BotFlex, which is a network-based tool for botnet detection, composed by a Complex Event Processing (CEP) engine and on a correlation framework that continuously receives events and correlates them according to rules. BotFlex's results are compared to BotHunter \cite{gu2007bothunter}, but the evaluation relies in a own and not public data set.

% justificative for the dataset and present who else use it
Several approaches for network attack detections uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD dataset maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The autors recommend that none of the botnet families used in the training and crossvalidation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training. In contrast, we propose to evaluate each scenario of CTU-13 individually, in a semi-supervised approach that does not rely on training data with anomalies.

@@

% Anomaly detection by subspace larning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of normal and anomalies according to principal components (normal) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

% Anomaly detection by robust subspace larning
Robust subspace learning has receiving a growing atention of researchers aiming the development of anomlay detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the problem of decomposing a given data matrix $\textbf{M}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for robust a subspace learning review.

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adoted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}.

Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on robust feature selection and robust PCA for network traffic anomaly detection. It is an unsupervised learning algorithm that allows matching the specificities of the traffic the detector is operating on, which can vary widely both in terms of the most frequent attacks and the relative weight of licit applications.

Our robust PCA detector does not require having a perfect ground-truth for training; it just trains from the (eventually contaminated) background traffic. It is, in fact, an unsupervised learning algorithm. Moreover, the addition of feature selection allows matching the specificities of the traffic the detector is operating on, which can vary widely both in terms of the most frequent attacks and the relative weight of licit applications. Feature selection does require training with a labeled data set bu. One important strength of our anomaly detector is its adaptability to different network environments and traffic conditions

@@

Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

@@
